Optimizing Abstract Abstract Machines

### 2012-11-13 paper/paper.tex
More work on Glance section.

\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}

\begin{figure}
\hrule
\vspace{2mm}
\begin{alltt}
(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
(define (((mult m1) m2) mf) (m2 (m1 mf)))
(define ((((plus p1) p2) pf) x)
  ((p1 pf) ((p2 pf) x)))

@@

@@

\begin{figure}
\hrule
\vspace{2mm}
\begin{alltt}
(define (id x) x)
(define (f n)
  (if (<= n 1) 1 (* n (f (- n 1)))))
(define (g n)
  (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
(print (+ ((id f) 3) ((id g) 4)))
\end{alltt}
\hrule
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=4in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.7in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\end{figure}

\section{An abstract$^{\text 2}$ machine for ISWIM}
\label{sec:aam}

>>>>

\newcommand{\church}[1]{\(\ulcorner{\tt #1}\urcorner\)}
\begin{figure}
\hrule
\vspace{2mm}
\begin{alltt}
(define (zero? n) ((n (\(\lambda\) (zx) #f)) #t))
(define (((mult m1) m2) mf) (m2 (m1 mf)))
(define ((((plus p1) p2) pf) x)
  ((p1 pf) ((p2 pf) x)))

@@

\begin{figure}[t]
\hrule
\vspace{2mm}
\begin{alltt}
(define (id x) x)
(define (f n)
  (if (<= n 1) 1 (* n (f (- n 1)))))
(define (g n)
  (if (<= n 1) 1 (+ (* n n) (g (- n 1)))))
(print (+ ((id f) 3) ((id g) 4)))
\end{alltt}
\hrule
\begin{center}
\begin{tabular}{ccc}
\raisebox{1ex-\height}{
\includegraphics[height=4in]{introspective-base.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=3.7in]{introspective-lazy.pdf}}
&
\raisebox{1ex-\height}{
\includegraphics[height=2.2in]{introspective-lazyc.pdf}}
\\
(a) Baseline
&
(b) Lazy
&
(c) Compiled (\& lazy)
\end{tabular}
\end{center}
\caption{Example state graphs for the program above.  Part (a) shows
  the result of the baseline analyzer.  It has long ``corridor''
  transitions and ``diamond'' subgraphs that fan-out from
  nondeterminism and fan-in from joins.  Part (b) shows the result of
  performing nondeterminism lazily and thus avoids many of the diamond
  subgraphs.  Part (c) shows the result of abstract compilation that
  removes intepretive overhead in the form of intermediate states,
  thus minimizing the corridor transitions.  The end result is a more
  compact abstraction of the program that can be generated faster.}
\label{fig:state-graphs}
\end{figure}

The AAM approach, in essence, does the following: it takes a
machine-based view of computation and turns it into a \emph{finitary
  approximation} by bounding the size of the store.  With a limited
address space, the store must map addresses to \emph{sets} of values.
Store updates are interpreted as joins, and store dereferences are
interpreted by non-deterministic choice of an element from a set.  The
result of analyzing a program is a finite directed graph where nodes
in the graph are (abstract) machine states and edges denote machine
transitions between states.

The techniques we propose for optimizing analysis fall into the
following categories:
\begin{enumerate}
\item generate fewer states by avoiding the eager exploration of
  non-deterministic choices that will later collapse into a single
  join point.  We accompish this by applying lazy evaluation
  techniques so that non-determinism is evaluated \emph{by need}.

\item generate fewer states by avoiding unnecessary, intermediate
  states of a computation.  We accomplish this by applying compilation
  techniques from functional languages to avoid interpretive overhead
  in the machine transition system.

\item generate states faster.  We accomplish this by better algorithm
  design in the fixpoint computation we use to generate state graphs.
\end{enumerate}
Figure~\ref{fig:state-graphs} shows the effect of (1) and (2) for a
small example due to Earl, et
al.~\cite{dvanhorn:Earl2012Introspective}.
By generating significantly fewer states at a significantly faster
rate, we are able to achieve large performance improvements in terms
of both time and space.

@@

\newpage
\section{An abstract$^{\text 2}$ machine for ISWIM}
\label{sec:aam}
